---
title: "TARGET Osteosarcoma Copy Number Survival Example"
author: "Sean Davis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TARGET Osteosarcoma Copy Number Survival Example}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r results='hide', echo=FALSE}
library(knitr)
opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)
```

# Installation

```{r eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite('seandavi/TargetOsteoAnalysis')
```

```{r}
library(TargetOsteoAnalysis)
```

# Load copy number data

The copy number data can be easily loaded, including the 
clinical information, with a single command.

```{r}
cn_se = target_gistic_se()
cn_se
```

# Survival analysis

Either overall survival or relapse-free survival can be easily 
tested for any gene, as the GISTIC data are gene-based.

The following code simply create the `Surv` objects that encode
the survival clinical data.

```{r}
library(survival)
library(SummarizedExperiment)
os_obj = Surv(time = colData(cn_se)$`Time to First Event in Days`, 
              event = colData(cn_se)$`Vital Status`=='Dead')
rfs_obj = Surv(time = colData(cn_se)$`Time to First Event in Days`, 
               event = !is.na(colData(cn_se)$`Time to first relapse in days`))
```

The next code block uses Cox Proportional Hazards to test the 
effect of the copy number of the _MYC_ gene on survival, controlling
for metastasis status at diagnosis.

```{r}
library(survival)
metastatic = !is.na(colData(cn_se)$`Metastasis site`)
cn = assay(cn_se, 'cn')[rowData(cn_se)$`Gene Symbol`=='MYC',]
res = coxph(os_obj ~ cn*metastatic)
res
```

As noted above, the copy number (`cn`) term is not statistically 
significant, but the metastatic status at diagnosis is highly significant.
