---
title: "ShatterSeq"
author: "Sean Davis"
date: "9/17/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)
library(TargetOsteoAnalysis)
ssres = shatterseq_results()
```

## Structural Variants per Sample

```{r}
library(ggplot2)
library(dplyr)
svs_per_sample = ssres %>% select(SampleName, number_SVs_sample) %>% unique()
hist(svs_per_sample$number_SVs_sample, main = "Number of SVs per sample", xlab="Number of SVs")
```

## Filtering criteria

- High confidence: at least 6 interleaved intrachromosomal SVs, 7 contiguous segments oscillating between 2 CN states, the fragment joins test, and either the chromosomal enrichment
or the exponential distribution of breakpoints test.
- High confidence: at least 3 interleaved intrachromosomal SVs and 4 or more interchromosomal
SVs, 7 contiguous segments oscillating between 2 CN states and the fragment joins test.

## Filtering code

```{r, echo=TRUE}
ssres2= ssres %>% mutate(high_conf = ((number_intra_SV>5 & 
                       max_number_oscillating_CN_segments_2_states>6 &
                       pval_fragment_joins<0.01 & 
                         (chr_breakpoint_enrichment<0.01 | pval_exp_chr<0.01))) 
                       |
                         ((number_intra_SV>3 & (inter_number_DEL+inter_number_h2hINV+inter_number_t2tINV+inter_number_DUP>4)) 
                          & pval_fragment_joins<0.01 
                          & max_number_oscillating_CN_segments_2_states>6))
```

## Per sample results


```{r}
tmp = ssres2 %>% filter(high_conf==TRUE) %>% select(SampleName, chrom)
tmp = tmp %>% select(SampleName) %>% group_by(SampleName) %>% summarize(chromothripsis_chroms=n())
ggplot(tmp, aes(x=SampleName, y=chromothripsis_chroms)) + geom_bar(stat = 'identity') + theme(axis.text.x=element_text(angle = -45, hjust = 0))
```


## Per chromosome results


```{r}
tmp = ssres2 %>% filter(high_conf==TRUE) %>% select(SampleName, chrom)
tmp = tmp %>% select(chrom) %>% group_by(chrom) %>% summarize(chromothripsis_samples=n())
ggplot(tmp, aes(x=chrom, y=chromothripsis_samples)) + geom_bar(stat = 'identity')
```
